---
title: "Matching data from two independent surveys: labour force and living conditions"
author: "Ines Garmendia"
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{micromatch package}
  %\usepackage[utf8]{inputenc}
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{micromatch package: Example}
-->

```{r, echo=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
        if (!is.null(n <- options$out.lines)) {
                x = unlist(stringr::str_split(x, '\n'))
                if (length(x) > n) {
                        # truncate the output
                        x = c(head(x, n), '....\n')
                }
                x = paste(x, collapse = '\n') # paste first n lines together
        }
        hook_output(x, options)
})
opts_chunk$set(out.lines = 10)
```


About `micromatch`
==================

`micromatch` provides a set of utilities to ease the task of statistically matching microdata files from official statistics.

For a general overview, please refer to the package vignette.

About this document
===================

This documents presents the use of `micromatch` package through a real matching
example with two independent surveys from Eustat, the Basque Statistical Office.

The example: matching `ecv` and `pra`
===========================================

`micromatch` includes data from two separate surveys conducted by Eustat (The Basque Statistical Office) during the 4th quarter of 2009: 

1. the Labour Force Survey (Encuesta de Población en Relación con la Actividad, `pra`), and 
2. the Living Conditions Survey (Encuesta de Condiciones de Vida, `ecv`)

(Please refer to the package documentation for a full description of these datasets).

In the following steps, we will see how a synthetic file can be obtained starting from these two separate data frames, `ecv` and `pra`. Each step of the matching task (Fig 2) is covered by a set of functions in `micromatch`.

### Step 1: Specify the purpose of matching

First of the main elements of the matching task need to be specified:

* The list of _shared variables_ (i.e. the variables common to both files. Typically: age, sex, education level...)
* A list of _specific variables_ for the first file, A
* A list of _specific variables_ for the second file, B
* Optionally, some _stratum variables_, i.e. variables defining separate sub-groups on the population (typically sex or age)
* Possibly, some _weight variables_, i.e. variables used for estimating values for all the population)

`ecv` gives a bunch of living condition measures such as frequency of social relations, economic status, health problems, and so on. `pra` focuses on the labour market and produces a segmentation of the population in 5 categories. In this example the aim is to obtain contingency tables that cross each of the `ecv` items with the labour market segmentation provieded by `pra`.

Because `pra` has more than twice as observations than `ecv`, `pra` will act as donor and `ecv` as recipient. This means that the unique specific variable in `pra` (i.e. the labour market segmentation) will be donated (i.e. imputed) into the `ecv` file. This will produce a synthetic file with one additional variable (column) in the initial `ecv` file.

The specific variable in `pra` is the labour market segmentation (variable `labour`) with the following categories:

```{r, warning=FALSE, message=FALSE}
library(micromatch)
data(pra)
levels(pra$labour)
```

Since `ecv` is a quite long questionnaire measuring many different aspects of quality and style of life, a sample of interesting items was selected for inclusion in this package:

```{r}
data(ecv)
str(ecv[,13:25])
```

### Step 2: Select matching variables

The shared variables between `ecv` and `pra` are shown in Table 1:

Variable   | Short Name | Type
------------- | ----- | -------------
Territory  | territory | 3 categories
Age  | age | 6 categories
Sex  | sex | 2 categories
Agesex  | agesex | 12 categories (combination of age and sex)
Family size  | famsize | 3 categories
Student?  | student | TRUE, FALSE
Seeking job?  | seekjob | TRUE, FALSE
Employed?  | employed | TRUE, FALSE
Unemployed?  | unemployed | TRUE, FALSE
Inactive?  | inactive | TRUE, FALSE
Hours at work  | workhours | 4 categories
Dedication to housework  | housework | 2 categories

**Table 1: Shared variables between `ecv` and `pra`**

        * Important Note: The data frames have been pre-processed in order to provide 
        the same names for the common variables, and the same category levels for the
        categorical values. These preprocessing task will have to be tackled prior to 
        using `micromatch`.

The shared variables have been unified:

```{r}
# In the first data frame: pra
str(pra[,1:12])
# In the second data frame: ecv
str(ecv[,1:12])
```

Each data frame has its weight variable, named `weights`.

For convenience, we will specify character vectors for each type of variable, as follows:

```{r defineVariables}
# Shared variables
varshared <- c("territory", "sex", "age", "agesex",
            "famsize", "student", "seekjob", "employed",
            "unemployed", "inactive", "workhours", "housework")
# Specific variables in the 1st file
varesp_A <- "labour"
# Specific variables in the 2nd file
varesp_B <- c("healthproblems", "languages", "holidaydest","sparetime",
              "indsocial", "famsocial", "equipment", "housemode",
              "ownvehicles", "ambientalcond", "econstatus", "income","moneyend")
# Weight variables (named the same way for both files)
weights <- "weights"
```

Now we proceed to define the recipient and donor files with all the previous variables:

```{r createObjects}
# Create recipient and donor files
# Donor: pra
don <- donor(data = pra, matchvars = varshared, specvars = varesp_A, weights = weights)
# Receptor: ecv
rec <- receptor(data = ecv, matchvars = varshared, specvars = varesp_B, weights = weights)
```

Now we will manipulate these objets in order to we reach a final definition for the matching task. More specifically: 

* we will select an optimal subset of `varshared`, which will act as the common variables (thus we will update the character vector for `mathcvars`), and 

* we will introduce stratum (i.e. group) variables: `stratavars`.

The common variables selected for matching must meet two conditions:

1. **Coherence**: The selected variables must show the same information across the files. The coherence refers both comparable definitions (i.e. wording of items in questionnaires) and to comparable empirical (observed) distributions.

2. **Predictive value**: The selected variables must be highly related to the specific variables in each of the files.

If these two conditions are met, then statistical matching will produce results which are (at least) coherent with the original information sources. 

(Please refer to the chapter on Validation to see what other conditions must be met in order to produce high quality results from a broader point of view.)

#### Step 2-1: Compare empirical distributions

First we need to discard variables whose empirical distributions are not comparable between the files. (Note that lack of equivalence with respect to the definition of concepts i.e., wording of items in the questionnaires, is also an important criterion in this step). 

We can use the `tabulate2cat` function to tabulate a single categorical variable in both files. Using `weights` is optional, and we can choose absolute or relative cell values with the argument `cell_values`:

```{r tabulateExample}
# Tabulate "territory", using weights and with absolute cell values
tabulate2cat(data_A = ecv, data_B = pra, var_A = "territory", var_B = "territory", 
             weights_A = "weights", weights_B = "weights", cell_values = "abs")
# Proportions (i.e. cell values relative to totals)
tabulate2cat(data_A = ecv, data_B = pra, var_A = "territory", var_B = "territory", 
             weights_A = "weights", weights_B = "weights", cell_values = "rel")
```

`plot2cat` function produces a barplot for a single categorical variable. As before, `weights` is optional, and we can choose absolute or relative cell values with the argument `cell_values`:

```{r plotExample}
# Plot "territory", using weights and with relative cell values (i.e. proportions)
plot2cat(data_A = ecv, data_B = pra, var_A = "territory", var_B = "territory", 
             weights_A = "weights", weights_B = "weights", cell_values = "rel")
```

Finally `similarity2cat` produces disimilarity/similarity measures for empirical distributions based on `StatMatch` package. Again, `weights` is optional:

```{r measuresExample}
# Plot "territory", using weights and with relative cell values (i.e. proportions)
similarity2cat(data_A = ecv, data_B = pra, var_A = "territory", var_B = "territory", 
             weights_A = "weights", weights_B = "weights")
```

Doing this on a variable-by-variable basis can be time-consuming; for this reason, specific methods have been included in `micromatch` that handle all variables defined as `matchvars` (i.e. common variables betwen the files) at once. Here is an illustration:

**All common variables at once**

* Tables.

```{r compareMatchvarsTables, eval=TRUE}
# Tables: proportions and weights
compare_matchvars(x = rec, y = don, cell_values = "rel", weights = TRUE)
```

* Plots. 

```{r compareMatchvarsPlots, eval=TRUE}
# Plots: proportions and weights
compare_matchvars(x = rec, y = don, type = "plot", cell_values = "rel", 
                  weights = TRUE)
```

* Disimilarity measures.

```{r compareMatchvarsMeasures, eval=TRUE}
# Measures: weights
compare_matchvars(x = rec, y = don, type = "measures", weights = TRUE)
```

Globally, we suspect there could be some type of incoherence in variables `famsize` (category 1 person), `unemployed` and `housework`. In any case, coherence between variables should be investigated further.

In the `ecv`-`pra` example, age and sex groups are determinant both for the labour market segmentation and for living conditions. The combined variable, `agesex`, has been created to include as a group/strata variable. 

We now remove variable `age`, `sex` and `agesex` from the common variables and include it as `stratavars` variable. `micromatch` has specific functions for easy removal (`remove`) and inclusion (`include`) of variables:

```{r}
rec1 <- remove(x = rec, vars = c("age", "sex", "agesex")) 
#removes variables and creates new object
rec2 <- include(x = rec1, vars = "agesex", as = "stratavars") 
#includes 'agesex' as strata
##
don1 <- remove(x = don, vars = c("age", "sex", "agesex"))
don2 <- include(don1, vars = "agesex", as = "stratavars")
```

Now we will repeat the analyses with objects `rec2` and `don2`, which include `agesex` as stratum variables. By setting option `strata` to `TRUE` in `compare_matchvars` method we get the same answers as before, by each stratum value (12 groups of age and sex, in this case):

**Results by `agesex` groups**

* Tables

```{r compareMatchStrataTables, eval=FALSE}
# Tables: proportions and weights
compare_matchvars(x = rec2, y = don2, type = "table", weights = TRUE, 
                  cell_values = "rel", strata = TRUE)
```

(Not printed)

* Plots

```{r compareMatchStrataPlots, eval= FALSE}
# Plots: proportions and weights
compare_matchvars(x = rec2, y = don2, type = "plot", weights = TRUE, strata = TRUE, 
                  cell_values = "rel")
```

(Not printed)

* Measures

```{r compareMatchStrataMeasures, eval= TRUE}
# Measures: weights
#compare_matchvars(x = rec2, y = don2, type = "measures", weights = TRUE, strata = TRUE)
#da error, porque hay casos con 0 casos en el estrato +65
compare_matchvars(x = remove(rec2, vars=c("unemployed", "employed", "inactive", "seekjob")), 
                  y = remove(don2, vars=c("unemployed", "employed", "inactive", "seekjob")),
                  type = "measures", weights = TRUE, strata = TRUE)
```

**First selection**

A first, tentative selection of variables by `agesex` values is done in table 2:

Stratum   | Selected variables
------------- | ------------------
Male.15-24  | student, seekjob
Female.15-24  | student, seekjob
Male.25-34  | seekjob, housework  
Female.25-34  | unemployed, housework
Male.35-44  | unemployed, famsize
Female.35-44  | unemployed, occupied
Male.45-54  | unemployed, seekjob
Female.45-54  | unemployed, seekjob
Male.55-64  | inactive, seekjob
Female.55-64  | inactive
Male.65+  | famsize
Female.65+  | famsize

*Table 2. A tentative selection of shared variables based on coherence study*

#### Step 2-2: Assess predictive value

For the pre-selected variables, we will assess the predictive value with respect to the `specvars` in each of the files.

We can use `predictvalue` method to assess the predictive value of `matchvars` with respect to `specvars`:

* In `pra` file, the (unique) specific variable is `labour`:

```{r,eval=TRUE}
predictvalue(x = remove(don, "workhours")) 
```

        * Note. `workhours` is excluded since the original function in `StatMatch`, 
        `pw.assoc`, returns an error. This is because this variable only applies to 
        occupied people:
        
```{r}
table(pra$labour, pra$workhours)
```

* In `ecv` file, we have a set of `specvars`:

```{r, eval=TRUE}
predictvalue(x = remove(rec, "workhours"))
```

Now we should inspect what happens within levels of `agesex`, i.e. repeat the previous analysis by strata.

For the pre-selected variables (see Table 2), the `select` and `predictvalue` methods can be easily combined to obtain results for `agesex` groups:

**Example: strata Male.15-24**

We inspect the predictive value of preselected variables: `student`, `seekjob` within this group.

In `pra`:

```{r}
levels(ecv$agesex)
don2.M1524 <- select_strata(x = don2, value = "Male.(15,24]")
table(slot(don2.M1524, "data")[,"agesex"]) #ok
varshared[-c(6,7)] #variables to keep
#
predictvalue(x = remove(don2.M1524 , varshared[-c(6,7)]))
```

In this case the predictive value is high so we decide to keep both: `student` and `seekjob`.
TODO: PRINT TABLES.

Proceeding this way, we finally keep all variables in Table 2 and proceed to perform hot-deck imputation.

### Step 3: Nearest neighbour hot-deck imputation

Take first 10 rows and concatenate

```{r}
rec2.prueba <- select_observations(x = rec2, obs = 1:5)
don2.prueba <- select_observations(x = don2, obs = 1:5)
concat.prueba <- concatenate(x = rec2.prueba, y = don2.prueba)
concat.prueba
```

Match via hot-deck
```{r}
filled.rec2.prueba <- match.hotdeck(x = rec2.prueba, y = don2.prueba)
filled.rec2.prueba
```

Match obsevations for strata "Male.15-24".

**With all variables!**

* Match

```{r, eval=FALSE, echo=TRUE}
#Select strata
rec2.M1524 <- select_strata(x = rec2, value = "Male.(15,24]")
don2.M1524 <- select_strata(x = don2, value = "Male.(15,24]")
#Match
filled.rec2.M1524 <- match.hotdeck(x = rec2.M1524, y = don2.M1524)
```

* Validate observed vs imputed `labour` variable.

```{r, eval=FALSE, echo=TRUE}
plot2cat(data_A = slot(filled.rec2.M1524, "data"), 
             data_B = slot(don2.M1524, "data"),
             var_A = "labour",
             var_B = "labour",
             weights_A = "weights",
             weights_B = "weights",
             cell_values = "rel")
```

Bad results!

**With variable selection**

* Match

```{r, eval=FALSE, echo=TRUE}
#Match
varshared[-c(6,7)] #variables to keep
filled.rec2.M1524 <- match.hotdeck(x = remove(rec2.M1524, vars = varshared[-c(6,7)]),
                                   y = remove(don2.M1524, vars = varshared[-c(6,7)]))
```

* Validate

```{r, eval=FALSE, echo=TRUE}
plot2cat(data_A = slot(filled.rec2.M1524, "data"), 
             data_B = slot(don2.M1524, "data"),
             var_A = "labour",
             var_B = "labour",
             weights_A = "weights",
             weights_B = "weights",
             cell_values = "rel")
```

Results are much better with variable selection.

        *TODO. Programar matching por estratos identificando una lista tipo...

```{r}
#c("student", "seekjob") %in% varshared
#varselected <- list(c("student", "seekjob"), )
#luego lapply replicando lo hecho con el primer estrato.
#englobar los resultados en un unico objeto.
```

        *TODO.Validar los resultados con un objeto que englobe los 'fusedfile' 
        imputados de todos los estratos.Con esto, termina el ejemplo ecv-pra.

        *TODO. Siguiente ejemplo: pisa-talis y MICE sobre fichero concatenado.


